---
title: "MACHINE LEARNING"
author: "Anushya K"
date: "11/10/2020"
output: html_document
---

# PREDICTION ASSIGNMENT #

## Background ##

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Data ##

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har.

## Approach ##

Our outcome variable is classe, a factor variable with 5 levels. For this data set, “participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:

exactly according to the specification (Class A)

throwing the elbows to the front (Class B)

lifting the dumbbell only halfway (Class C)

lowering the dumbbell only halfway (Class D)

throwing the hips to the front (Class E)

Two models will be tested using decision tree and random forest. The model with the highest accuracy will be chosen as our final model.

## Cross-validation ##

Cross-validation will be performed by subsampling our training data set randomly without replacement into 2 subsamples: TrainTrainingSet data (75% of the original Training data set) and TestTrainingSet data (25%). Our models will be fitted on the TrainTrainingSet data set, and tested on the TestTrainingSet data. Once the most accurate model is choosen, it will be tested on the original Testing data set.

## Expected out-of-sample error ##

The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data. Accuracy is the proportion of correct classified observation over the total sample in the TestTrainingSet data set. Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.

Our outcome variable “classe” is a factor variable. We split the Training dataset into TrainTrainingSet and TestTrainingSet datasets.

Install packages and load the required libraries:

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(lattice)
library(caret)
library(e1071)
library(randomForest)
library(rpart)
library(rpart.plot)
```

load and clean data:

```{r}
# set the seed to make the data reproducible

set.seed(123)

# loading and cleaning data

trainingset <- read.csv("A:/R/datasets/pml-training.csv", na.strings=c("NA","#DIV/0!", ""))
testingset <- read.csv("A:/R/datasets/pml-testing.csv", na.strings=c("NA","#DIV/0!", ""))
```

```{r}
# Check dimensions for number of variables and number of observations

dim(trainingset)
dim(testingset)
```

```{r}
# delete columns with missing values

trainingset<-trainingset[,colSums(is.na(trainingset)) == 0]
testingset <-testingset[,colSums(is.na(testingset)) == 0]

# delete irrelevant variables

trainingset   <-trainingset[,-c(1:7)]
testingset <-testingset[,-c(1:7)]

# new dataset

dim(trainingset)
dim(testingset)
head(trainingset)
head(testingset)
```

```{r}
# partition the data so that 75% of the training dataset into training and the remaining 25% to testing

subsamples <- createDataPartition(y=trainingset$classe, p=0.75, list=FALSE)
subTraining <- trainingset[subsamples, ] 
subTesting <- trainingset[-subsamples, ]
dim(subTraining)
dim(subTesting)
head(subTraining)
head(subTesting)
```

```{r}
# The variable "classe" contains 5 levels: A, B, C, D and E. A plot of the outcome variable will allow us to see the frequency of each levels in the TrainTrainingSet data set and compare one another.

plot(subTraining$classe, col="blue", main="Bar Plot of levels of the variable classe within the subTraining data set", xlab="classe levels", ylab="Frequency", ylim = c(0,5000))

```

Based on the graph above, we can see that each level frequency is within the same order of magnitude of each other. Level A is the most frequent while level D is the least frequent.

## Prediction Model 1: Decision Tree ##

```{r}

model1 <- rpart(classe ~ ., data=subTraining, method="class")
rpart.plot(model1, main = "classification tree", extra = 102, under = TRUE, faclen = 0)

```
```{r}

# Predicting:
prediction1 <- predict(model1, subTesting, type = "class")


subTesting$classe <- as.factor(subTesting$classe)


# Test results on our TestTrainingSet data set:
confusionMatrix(subTesting$classe, prediction1)
```

## prediction Model 2: Random Forest ##

```{r}
RF1 <- trainControl(method = "cv", number = 3, verboseIter = FALSE)


model2 <- randomForest(classe ~. , data=subTraining, method="rf")

# Predicting:
prediction2 <- predict(model2, subTesting, type = "class")

# Test results on subTesting data set:
confusionMatrix(prediction2, subTesting$classe)
```

## Decision on which Prediction Model to Use: ##

Random Forest algorithm performed better than Decision Trees. Accuracy for Random Forest model was 0.995 (95% CI: (0.993, 0.997)) compared to Decision Tree model with 0.739 (95% CI: (0.727, 0.752)). The Random Forests model is choosen. The expected out-of-sample error is estimated at 0.005, or 0.5%.

## Submission ##

Here is the final outcome based on the Prediction Model 2 (Random Forest) applied against the Testing dataset

```{r}
# predict outcome levels on the original Testing data set using Random Forest algorithm

predictfinal <- predict(model2, testingset, type="class")
predictfinal
```

```{r}
# Write files for submission

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictfinal)
```